---
title: "Session 3 Tutorial"
author: "Ahmed Aquib"
date: "2024-04-25"
output: html_document
---

# Session 3 Tutorial

## Introduction
This tutorial covers the basics of creating heatmaps using the `pheatmap` package in R. We'll start with a simple heatmap and gradually add more features and customizations.

## Prerequisites
Before we begin, make sure you have the `pheatmap` and `RColorBrewer` libraries installed and loaded:

```{r setup, include=FALSE}
library(pheatmap)
library(RColorBrewer)
```

## Simplest Heatmap
Let's start with the simplest heatmap:
```{r}
# Generate a 3x3 matrix with numbers 1 to 9
data_minimal <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)

# Create a minimalist heatmap
pheatmap(data_minimal)
```
## Adding Annotations
Now, let's add annotations to our heatmap:

```{r}
dim <- list(row=c(1,2,3), col=c("A","B","C"))
data_minimal <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE, dimnames = dim)
# Heatmap with annotations
pheatmap(data_minimal, display_numbers = TRUE, fontsize = 20)
```

By default, pheatmap applies hierarchical clustering to both rows and columns. To disable clustering, use the `cluster_rows = FALSE, cluster_cols = FALSE`

## Scaling Rows

```{r}
pheatmap(data_minimal, scale = "row", display_numbers = TRUE, fontsize = 20)
```

### Question: But what does scaling actually do??
```{r}
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

t(apply(data_minimal, 1, cal_z_score))
```
## A more complex Heatmap

```{r}
# Generate random data
data <- matrix(rnorm(150), 25, 6)
colnames(data) <- paste0("Sample_", 1:6)
rownames(data) <- paste0("Gene_", 1:25)

# Display the first few rows of the data
head(data)
```

## Customizing Heatmaps

```{r}
pheatmap(data,
         fontsize_col = 10,
         fontsize_row = 10,
         display_numbers = TRUE,
         number_color = "black", 
         fontsize_number = 6,
         border_color = "black")
```

## Customizing Cell Colors

```{r}
pheatmap(data,
         fontsize_col = 10,
         fontsize_row = 10,
         display_numbers = TRUE,
         number_color = "black", 
         fontsize_number = 6,
         border_color = "black",
         col = brewer.pal(10, 'RdYlGn'))
```

## Adding Annotations

```{r}
ann_df <- data.frame(Group = rep(c("Disease", "Control"), c(3, 3)),
                          Lymphocyte_count = rnorm(6))

row.names(ann_df) <- colnames(data)

gene_functions_df <- data.frame(gene_functions = rep(c('Oxidative_phosphorylation', 
                                                       'Cell_cycle',
                                                       'Immune_regulation',
                                                       'Signal_transduction',
                                                       'Transcription'), each=5))
row.names(gene_functions_df) <- rownames(data)

ann_colors <- list(
  gene_functions = c("Oxidative_phosphorylation" = "#F46D43",
              "Cell_cycle" = "#708238",
              "Immune_regulation" = "#9E0142",
              "Signal_transduction" = "beige", 
              "Transcription" = "violet"), 
  Group = c("Disease" = "darkgreen",
              "Control" = "blueviolet"),
  Lymphocyte_count = brewer.pal(5, 'PuBu')
)

pheatmap(data, 
         col = brewer.pal(10, 'RdYlGn'),
         #Adding Annotation Information
         annotation_row = gene_functions_df, 
         annotation_col = ann_df, 
         annotation_colors = ann_colors,
         # The same Customization
         fontsize_col = 10,
         fontsize_row = 10,
         display_numbers = TRUE,
         number_color = "black", 
         fontsize_number = 6,
         border_color = "black",
         #A heading
         main = "A Heatmap with Annotations") 

```

