---
title: "Session_2"
author: "Ahmed Aquib"
date: "2024-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

## Series On: Data Wrangling and Visualization in R

## Session 1: Plotting Data with Base R and Introduction to ggplot2

# Plotting Simple Data with Base R

## Histograms

Histograms show a distribution of a single variable continuous data.

```{r}
normal_data <- rnorm(200, mean=100, sd=20)
head(normal_data)
hist(normal_data)
```

## Scatterplots

When you want to check relationship between two variables.

```{r}
iris
head(iris)

plot(iris$Sepal.Length, iris$Sepal.Width)
```

```{r}
p <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width))+
    geom_point(aes(color = Species, shape = Species))+
    scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))
print(p)
```

## Barplots and Boxplots

```{r}
data("iris")
barplot(colMeans(iris[,1:4]))
boxplot(iris[,1:4])
```

# Let's get started with ggplot2

## Histograms

```{r}
# ggplot2 requires a tidy data_frame
df <- data.frame(var=normal_data)
ggplot(df)
```

```{r}
df <- data.frame(var=normal_data)
ggplot(df)+
  geom_histogram(aes(var), binwidth = 5, color="white", fill="azure4")
```

```{r}
df <- data.frame(var=normal_data)
ggplot(df)+
  geom_histogram(aes(var), binwidth = 5, color="white", fill="azure4") +
  labs(x="Variable",y="Frequency",title="Histogram of a Normal Discribution")
```

```{r}
# Define the dataframe
data <- data.frame(
  Sample_Name = c("GSM1275863", "GSM1275867", "GSM1275871", "GSM1275875", "GSM1275862", "GSM1275866", "GSM1275870", "GSM1275874"),
  Treatment = c("Dexamethasone", "Dexamethasone", "Dexamethasone", "Dexamethasone", "None", "None", "None", "None"),
  Number_of_Transcripts = c(10768, 10051, 11658, 10900, 11177, 11526, 11425, 11000),
  Total_Counts = c(18783120, 15144524, 30776089, 21135511, 20608402, 25311320, 24411867, 19094104)
)
```

## Scatterplots

```{r}
ggplot(data)
```

Let's add one single aesthetics to define our x and y of the plot

```{r}
ggplot(data) +
  geom_point(aes(x= Number_of_Transcripts, y = Total_Counts))
```

```{r}
ggplot(data) +
  geom_point(aes(x= Number_of_Transcripts, y = Total_Counts, color=Treatment))
```

```{r}
ggplot(data) +
  geom_point(aes(x= Number_of_Transcripts, y = Total_Counts, color=Treatment)) +
  scale_color_manual(values=c("red","black"),labels=c('Treated',"Untreated"))
```

```{r}
ggplot(data) +
  geom_point(aes(x= Number_of_Transcripts, y = Total_Counts, color=Treatment)) +
  scale_color_manual(values=c("red","black"),labels=c('Treated',"Untreated")) +
  geom_text(aes(x = Number_of_Transcripts, y = Total_Counts, label = Sample_Name), 
            vjust = -0.1, hjust = -0.1, size = 3)+
  xlim(10000,12000)
```

## Barplots

```{r}
summarized_data <- data %>% group_by(Treatment) %>% summarize(Mean=mean(Number_of_Transcripts), SD = sd(Number_of_Transcripts))
  
ggplot(summarized_data) +
  geom_bar(aes(x=Treatment,y=Mean), stat = "identity")
```

```{r}
ggplot(summarized_data,aes(x=Treatment,y=Mean, fill=Treatment)) +
  geom_bar(stat = "identity", width=0.6, position='identity') +
  geom_errorbar(aes(ymin=Mean-SD,ymax=Mean+SD), width=0.2, color="darkslategrey", size=1)
```

```{r}
ggplot(summarized_data,aes(x=Treatment,y=Mean, fill=Treatment)) +
    geom_bar(stat = "identity", width=0.6, position='identity',color="black") +
    geom_errorbar(aes(ymin=Mean-SD,ymax=Mean+SD), width=0.1, color="darkslategrey",size=1)+
     geom_point(data = data, aes(x = Treatment, y = Number_of_Transcripts, shape=Treatment), position = position_jitter(width = 0.2), size = 3, color="darkslategrey")
```

## Boxplots

```{r}
ggplot(data, aes(x = Treatment, y = Number_of_Transcripts, fill = Treatment)) +
  geom_boxplot(color = "black", alpha = 0.8) +  # Add boxplot with border color and transparency
  geom_jitter(width = 0.2, size = 3, alpha = 0.7) +  # Add jittered points with size and transparency
  scale_fill_manual(values = c("lightcoral", "lightblue4"), labels = c('Treated', "Untreated")) +  # Custom color scheme
  labs(
    x = "Treatment", y = "Number of Transcripts",  # Axis labels
    title = "Number of Transcripts by Treatment",  # Plot title
    subtitle = "Boxplot"  # Plot subtitle
  ) 
```
